Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-09-21 manuscript/manuscript.tex
Minor edits after suggestions by Santiago

To take into account the curvature of the Earth, many global forward modelling methods
are defined in geocentric spherical coordinates.
A common approach is to discretize the Earth into tesseroids (spherical prisms),
which are defined by pairs of latitude, longitude, and
radial boundaries (see Fig.~\ref{fig:tesseroid}).
The gravitational fields generated by an arbitrary
tesseroid on any external point are given by volume
integrals that are generally approximated by numerically.
The literature offers two main approaches: one involves Taylor series expansion
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre
Quadrature (GLQ)
\citep{Asgharzadeh2007, Wild-Pfeiffer2008, Li2011, Uieda2016}.
The Taylor series expansion is not well suited to develop an algorithm for
a density varying with depth according to an arbitrary function.
Different series expansion terms would have to obtained for each density function
desired.
Conversely, an arbitrary density function can be included in the GLQ without
any change to the integration method.

@@

\begin{table}
\caption{
    Description of computation grids used to characterize the accuracy of
    the numerical integration.
    Every grid consist of a set of $10 \times 10$ points.
}
\label{tab:grids}
\begin{tabular}{lccc}
    Name & Grid region (degrees) & Grid height (km)
    \\ \hline
    Pole & $0\text{E}/1\text{E}/89\text{N}/90\text{N}$ & 2 \\
    Equator & $0\text{E}/1\text{E}/0\text{N}/1\text{N}$ & 2 \\
    Satellite & $0\text{E}/1\text{E}/89\text{N}/90\text{N}$ & 260 \\
    Big Grid & $0\text{E}/30\text{E}/60\text{N}/90\text{N}$ & 2 \\
\end{tabular}
\end{table}

@@

In order to compare the numerical results with the analytical solution we
must build a spherical shell made of tesseroids.
We use a single radial layer of $30^\circ \times 30^\circ$ tesseroids to
build the spherical shell.
We compute their gravitational effects on $10 \times 10$ point grids at
different heights and locations.
These grids are defined in Table \ref{tab:grids}.
The computation of the gravitational effects on each grid is repeated for
two shells with thickness 1km and 35 km.
The outer radius of both shells are equal to the mean Earth radius and
the density varies linearly (Eq. \ref{eq:density-linear})
with angular coefficient

@@

We compute the gravitational potential ($V$), the vertical component of the gradient
($g_z$), and the diagonal components of the Marussi tensor ($g_{xx}$, $g_{yy}$,
$g_{zz}$) for each grid in Table \ref{tab:grids}.
Other components are equal to zero outside of the shell and are thus omitted from the
analysis.
The computations are repeated for values of $D$ ranging from 0.5 to 10 with a step of
0.5.
We then will calculate the maximum absolute difference between these results and the
analytical solutions derived in Appendix \ref{sec:shell}.
The differences are shown in Fig. \ref{fig:D-linear} as relative to the spherical shell
values.
We omit the differences for $g_{xx}$ and $g_{yy}$ because they are equivalent to the
ones obtained for $g_{zz}$.
Finally, we set the optimal value of $D$ as the minimum value at which the corresponding
error of the numerical approximation is lower than 0.1\%.

@@

Because there is no direct relation between the values assigned to $D$ and $\delta$ and
the error of the computation, we empirically determined the optimum values for these
parameters.
The values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimum values were a linear and an
exponential function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
We analyzed the error for exponential functions ranging from smooth to sharp variations
in depth to test the accuracy of the density-based discretization.
Through this analysis, we have obtained optimal values for the distance-size ratio of
1, 2 and 8 for the potential, its gradient, and the Marussi tensor components,
respectively.
These values are in agreement with previous results for a homogeneous density tesseroid
\citep{Uieda2016}.
The optimum $\delta$ values are 0.2 for the potential and its gradient components, while
a $\delta = 0.01$ must be used for the Marussi tensor components in order to ensure a
0.1\% precision.
A wide range of exponential functions were used to infer these results and they are
conservative estimates, chosen to prioritize accuracy over speed of computations.
Therefore, they can be safely generalized to other common depth-density relations, like
quadratic functions, without loss of accuracy.

>>>>

To take into account the curvature of the Earth, many global forward modelling methods
are defined in geocentric spherical coordinates.
A common approach is to discretize the Earth into tesseroids (spherical prisms),
which are defined by pairs of latitude, longitude, and
radial boundaries (see Fig.~\ref{fig:tesseroid}).
The gravitational fields generated by an arbitrary
tesseroid on any external point are given by volume
integrals that must be numerically approximated.
The literature offers two main approaches: one involves Taylor series expansion
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre
Quadrature (GLQ)
\citep{Asgharzadeh2007, Wild-Pfeiffer2008, Li2011, Uieda2016}.
The Taylor series expansion is not well suited to develop an algorithm for
a density varying with depth according to an arbitrary function.
Different series expansion terms would have to obtained for each density function
desired.
Conversely, an arbitrary density function can be included in the GLQ without
any change to the integration method.

@@

\begin{table}
\caption{
    Description of the computation grids used to characterize the accuracy of the
    numerical integration.
    All grids consist of a set of $10 \times 10$ points.
}
\label{tab:grids}
\begin{tabular}{lccc}
    Name & Grid size & Grid region (degrees) & Grid height (km)
    \\ \hline
    Pole & $1^\circ \times 1^\circ$ & $0\text{E}/1\text{E}/89\text{N}/90\text{N}$ & 2 \\
    Equator & $1^\circ \times 1^\circ$ & $0\text{E}/1\text{E}/0\text{N}/1\text{N}$ & 2 \\
    Satellite & $1^\circ \times 1^\circ$ & $0\text{E}/1\text{E}/89\text{N}/90\text{N}$ & 260 \\
    Big Grid & $30^\circ \times 30^\circ$ & $0\text{E}/30\text{E}/60\text{N}/90\text{N}$ & 2 \\
\end{tabular}
\end{table}

@@

In order to compare the numerical results with the analytical solution we
must build a spherical shell made of tesseroids.
The outer radius of the shell is equal to the mean Earth radius and we repeat the
calculations for shell thicknesses of 1km and 35km.
We discretize the shell into a single layer of $30^\circ \times 30^\circ$ tesseroids and
compute their gravitational effects on $10 \times 10$ point grids with different
resolutions, heights, and locations on the sphere.
These grids are defined in Table \ref{tab:grids}.
The density of the shell and tesseroid model varies linearly with $r$
(Eq. \ref{eq:density-linear}) with angular coefficient

@@

We compute the gravitational potential ($V$), the vertical component of the gradient
($g_z$), and the diagonal components of the Marussi tensor ($g_{xx}$, $g_{yy}$,
$g_{zz}$) for each grid in Table \ref{tab:grids}.
Other components are equal to zero outside of the shell and are thus omitted from the
analysis.
The computations are repeated for values of $D$ ranging from 0.5 to 10 with a step of
0.5.
We will then calculate the maximum absolute difference between these results and the
analytical solutions derived in Appendix \ref{sec:shell}.
The differences are shown in Fig. \ref{fig:D-linear} as relative to the spherical shell
values.
We omit the differences for $g_{xx}$ and $g_{yy}$ because they are equivalent to the
ones obtained for $g_{zz}$.
Finally, we set the optimal value of $D$ as the minimum value at which the corresponding
error of the numerical approximation is lower than 0.1\%.

@@

Because there is no direct relation between the values assigned to $D$ and $\delta$ and
the error of the computation, we empirically determined the optimum values for these
parameters.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimum values were a linear and an
exponential function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
We analyzed the error for exponential functions ranging from smooth to sharp variations
in depth to test the accuracy of the density-based discretization.
Through this analysis, we have obtained optimal values for the distance-size ratio of
1, 2 and 8 for the potential, its gradient, and the Marussi tensor components,
respectively.
These values are in agreement with previous results for a homogeneous density tesseroid
\citep{Uieda2016}.
The optimum $\delta$ values are 0.2 for the potential and its gradient components, while
a $\delta = 0.01$ must be used for the Marussi tensor components in order to ensure a
0.1\% precision.
A wide range of exponential functions were used to infer these results and they are
conservative estimates, chosen to prioritize accuracy over speed of computations.
Therefore, they can be safely generalized to other common depth-density relations, like
quadratic functions, without loss of accuracy.
